Pure Spin Current Injection in Hydrogenated Graphene Structures

### 2017-07-01 mu/article/article.tex
sab
Corrections completed

\usepackage{amsmath}    % need for subequations
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs 
\usepackage{blindtext}  % fill text
\usepackage[dvipsnames]{xcolor}
\usepackage{multirow}
% \usepackage{showframe}
% \usepackage[inline]{showlabels}

@@

{
\color{blue}
Spintronics is an emerging research field of electronics in which the
manipulation and transport of spin of electrons in a solid state 
media plays the determining role adding a new degree of freedom to the
conventional charge manipulation.\cite{wolfSC04,fabianAPS07}
% 
At present there is an increasing interest in attain the same level of control
over the transport of spin at micro or nano scales as has been done for the
flow of charge in typical electronic devices.\cite{awschalomNP2007} Some
semiconductor spintronics devices have been proposed \cite{majumdarAPL06,
dattaAPL90} and some of them require spin polarized electrical current
\cite{awschalomSSBM13} or pure spin currents (PSC). In PSCs there is no net
motion of charge; spin-up electrons move in a given direction while spin-down
electrons travel in the opposite one. This phenomena can result from spin
injection,\cite{malPRB03} Hall Effects,\cite{sinovaPRB04} interference of two
optical beams,\cite{bhatPRL00, najmaiePRB03} or one photon absorption of
linearly polarized light\cite{bhatPRL05} and has been observed in gallium
arsenide (GaAs),\cite{zhaoPRL2006, stevensPRL03} aluminum-gallium arsenide
(AlGaAs),\cite{stevensPRL03} and Co$_2$FeSi.\cite{kimuraNGPAM12}

% Since its discovery, graphene has attracted a great interest in materials
% science and condensed matter physics being a new class of materials that are
% only one atom thick\cite{geimNM07}. 
% 
Graphene, an allotrope of carbon with hexagonal 2D lattice structure presents
properties like fractional quantum Hall effect at room temperature, excellent
thermal transport properties, excellent conductivity\cite{heerscheNat07} and
strength \cite{geimNM07, reinaNL08, novoselov2S07, balandinNL08} being then a
perfect platform to be used in two-dimensions electronic systems; however most
electronic applications are disabled by the absence of a semiconducting gap.
Recent studies demonstrate that the band gap of graphene can be opened by
applying an electric field,\cite{zhangN09} reducing the surface
area,\cite{hanPRL07} or applying uniaxial strain.\cite{niACSN08} Another
possibility to open the gap is by doping; this has been successfully achieved
using nitrogen,\cite{weiNL2009} boron-nitrogen,\cite{guoIJ11}
silicon,\cite{colettiPRB10} noble-metals,\cite{varykhalovPRB10} and
hydrogen.\cite{eliasS09, guisingerNL09, samarakoonACSN10}
% 
Depending on the percentage of hydrogenation and spatial configurations of
hydrogen-carbon bonds, hydrogenated graphene can result in different spatial
configurations.
% 
In this paper we present two 50\% hydrogenated graphene non centrosymmetric
structures both presenting a discernible band gap: the \emph{up} structure,
shown in Fig. \ref{fig:up-struc}, has hydrogen atoms bonded to the carbon layer
only in the upper side of the structure while the \emph{alt} structure, shown
in Fig. \ref{fig:alt-struc}, has hydrogen alternating in the upper and bottom
sides of the carbon slab.\cite{zapataPSB2016}

Using those structures we address a theoretical study of the spin velocity
injection (SVI) by one-photon absorption of linearly polarized light that does
not seem to have been reported previously.
% 
We calculated the responses for the particular cases when the spin of electrons
is directed along the the $z$ Cartesian coordinate, perpendicular to the $xy$
plane of the structure, or for the case when the velocity is directed along the
$x$ and $y$ Cartesian direction on the $xy$ plane of the structures.
% 
The SVI ($\mathcal{V}^{\mathrm{ab}}(\omega,\alpha)$) is optical effect that
quantifies the velocity at which a PSC moves along the Cartesian direction
$\mathrm{b}$ with the spin of electron polarized along the Cartesian direction
$\mathrm{a}$. One photon absorption of linearly polarized light can promote a
distribution of electrons in $\mathbf{k}$ space regardless the symmetry of the
material resulting in a not net electrical current. Then, the electrons excited
to the conduction bands at opposite $\mathbf{k}$ points will result in opposite
spin polarizations producing no net spin injection.\cite{bhatPRL05} If the
crystalline structure of the material is not centrosymmetric the spin
polarization injected at a given $\mathbf{k}$ point could not
vanish\cite{alvaradoPRL85, schmiedeskampPRL88} and then a PSC will be produced
since the velocities of electrons at opposite $\mathbf{k}$ points are opposite.

This paper is organized as follows. In Section \ref{sec:theory} we present the
theory and formulas that describe PSC and SVI. In Section \ref{sec:results} we
describe the details of calculations and the corresponding SVI spectra for the 
\emph{up} and \emph{alt} structures. Finally, in Section

}

@@

In this section, we report a summary of the theory that involves the PSC
phenomena from which rises the SVI treated in this paper. The full description 
of it was presented by Bhat and Sipe et. al.\cite{bhatPRL05}

As mentioned before in Section \ref{sec:introduction}, In PSCs there is no net
motion of charge and spin-up electrons move in a given direction while spin-
down electrons travel in the opposite one. This effect can result from one
photon absorption of linearly polarized light by a semiconductor, with filled
valence bands and empty conduction bands, illuminated by light with photon
energy larger than the energy gap.
% note 1
Using i.e. a single weak continuous linearly polarized laser beam, is possible
to promote electrons in $\mathbf{k}$ space regardless the symmetry of the
system resulting in a net current equal to zero. Then, if the system presents
inversion of symmetry, electrons promoted to the conduction bands at opposite
$\mathbf{k}$ points will have opposite spin polarization resulting in a total
spin injection equal to zero. If the phenomena is produced in a
noncentrosymmetric semiconducting media the spin polarization injected at a
given $\mathbf{k}$ point can be held\cite{alvaradoPRL85}, resulting in a PSC
because the velocity of electrons at opposite $\mathbf{k}$ points are in
opposite directions.

@@

We define the SVI as the velocity at which the spin, polarized along the
direction $\mathrm{a}$, propagates along the direction $\mathrm{b}$ as
\begin{equation}\label{eq:vab-w}
\mathcal{V}^{\mathrm{ab}}(\omega) \equiv
\frac{\dot{K}^{\mathrm{ab}}(\omega)}{(\hbar/2) \dot{n}(\omega)},
\end{equation}  
where the pure spin density injection current, $\dot{K}^{\mathrm{ab}}
(\omega)$, and the carrier injection rate, $\dot{n}(\omega)$, are given by
% 
\begin{equation}\label{eq:dot-kn}
\begin{aligned}
\dot{K}^{\mathrm{ab}}(\omega) 
&=
\mu^{\mathrm{abcd}}(\omega)
E^{\mathrm{c}}(\omega) E^{\mathrm{d*}}(\omega),
% \label{eq:dotk} 
\\
\dot{n}(\omega)
&=
\xi^{\mathrm{ab}}(\omega) E^{c }(\omega) E^{d*}(\omega),
% \label{eq:dotn}
\end{aligned}
\end{equation}
where the roman superscripts denote Cartesian directions, and if repeated, they
are to be summed over, the $\xi^{\mathrm{ab}}(\omega)$ are the carrier
generation rate tensor components and $\mu^{\mathrm{abcd}}(\omega)$ are the
pure spin-current pseudotensor components given by\cite{bhatPRL05}
\begin{equation*}\label{eq:mu}
\begin{aligned}
\mu^{\mathrm{abcd}}  (\omega) &
=
\frac{\pi e^{2}}{\hbar^{2}} \int 
\frac{d^{3}K}{8 \pi^{3}} \times \\
\sum_{vcc'}^{'}
\mathrm{Re} & \left[ K^{\mathrm{ab}}_{cc'} 
\left( 
r^{\mathrm{c}}_{vc'} 
r^{\mathrm{d}}_{cv } +
(c \leftrightarrow d)
\right) 
\right] \delta(\omega-\omega_{cv}),
\end{aligned}
\end{equation*}
where $K^{\mathrm{ab}}_{mn}(\mathbf{k}) =
\sum_{\ell}v^{\mathrm{a}}_{nl}(\mathbf{k}) S^{\mathrm{b}}_{lm}(\mathbf{k})$ are
the spin current matrix elements that, using time reversal invariance, satisfy
the relation $K^{\mathrm{ab}}_{nm}(\mathbf{-k}) = K^{\mathrm{ab*}}(\mathbf{k})
$. 
% 
The $'$ in the sum means that $c$ and $c'$ are quasi degenerate states and the
sum only covers these states and since $\mu^{\mathrm{abcd}}(\omega)$ is real we
have that $\mu^{\mathrm{abcd}}(\omega) = \mu^{\mathrm{abdc}}(\omega)$.
% 
Since we have 2D structures we use an incoming electric field parallel to the
surface, $\mathbf{E}^{a}(\omega) = E^{\mathrm{a}} e^{i(\alpha + \omega t)}$,
where the angle $\alpha$ corresponds to the linear polarization angle measured
positively in the counter-clockwise from the $x$ direction on the
surface of the structures.
% 
Then from Eq. \eqref{eq:dot-kn} we can rewrite the Eq. \eqref{eq:vab-w} and 
including the polarization angle dependence we have
\begin{widetext}
\begin{align}
\mathcal{V}^{\mathrm{ab}}(\omega,\alpha)
&= \frac{2}{\hbar}
\frac{\mu^{\mathrm{abxx}}(\omega)
E^{2}(\omega)\cos^{2}(\alpha) + 
\mu^{\mathrm{abyy}}(\omega)
E^{2}(\omega)\sin^{2}(\alpha) + 
2\mu^{\mathrm{abxy}}(\omega)
E^{2}(\omega)\cos(\alpha)\sin(\alpha)}
{\xi^{\mathrm{xx}}(\omega)
E^{2}(\omega)\cos^{2}(\alpha) + 
\xi^{\mathrm{yy}}(\omega)
E^{2}(\omega)\sin^{2}(\alpha)},
\nonumber \\
&= \frac{2}{\hbar}
\frac{\mu^{\mathrm{abxx}}(\omega)\cos^{2}(\alpha) + 
\mu^{\mathrm{abyy}}(\omega)\sin^{2}(\alpha) + 
\mu^{\mathrm{abxy}}(\omega)\sin(2\alpha)}
{\xi^{\mathrm{xx}}(\omega)\cos^{2}(\alpha) + 
\xi^{\mathrm{yy}}(\omega)\sin^{2}(\alpha)},
\label{eq:vab-aw}
\end{align}
\end{widetext}
% 
If the polarization angle is fixed to $\alpha = \frac{\pi}{4}$ the previous
expression can be reduced to 
\begin{equation*}\label{eq:vab-90deg}
\mathcal{V}^{\mathrm{ab}} (\omega)
= \frac{2}{\hbar}
\frac{\mu^{\mathrm{abxx}}(\omega) + \mu^{\mathrm{abyy}}(\omega) + 
2\mu^{\mathrm{abxy}}(\omega)}
{\xi^{\mathrm{xx}}(\omega) + \xi^{\mathrm{yy}}(\omega)}.
\end{equation*}
Two interesting possibilities to analyze the SVI are fixing the spin along any
of the three Cartesian coordinates and, for our cases, particularly for the
spin directed along $z$ directed perpendicularly to the surface of the
structure; or fixing the velocity along the $x$ or $y$ Cartesian coordinates on
the $xy$ plane of the structures. Also we can analyze the SVI contribution
coming from each layer of the structure. In following subsections we present
these three cases.

@@

Analyzing the SVI, Eq. \eqref{eq:vab-aw}, we define the magnitude of the spin
velocity with spin polarized along the $\mathrm{b}$ Cartesian coordinate as
\begin{equation}
|\mathcal{V}_{\sigma^{\mathrm{b}}}(\omega,\alpha)| 
=
\sqrt{
[\mathcal{V}^{\mathrm{xb}}(\omega,\alpha)]^{2}\ +
[\mathcal{V}^{\mathrm{yb}}(\omega,\alpha)]^{2}\ 
}, 
\label{eq:vs-mag}
\end{equation}
and the angle at which the spin velocity is directed on the $xy$ plane as
\begin{equation}
\gamma_{\mathrm{b}} (\omega,\alpha)
=
\tan^{-1} \left( \frac{\mathcal{V}^{\mathrm{yb}}(\omega,\alpha)}
{\mathcal{V}^{\mathrm{xb}}(\omega,\alpha)} \right),
\label{eq:gamma-ang}
\end{equation}
where this angle is measured in the counter-clockwise direction from the
positive $x$ Cartesian coordinate. We also define two special
angles
\begin{align}
\gamma_{\mathrm{b \parallel}}(\omega,\alpha) &= \alpha, 
\label{eq:gamma-par} 
\\
\gamma_{\mathrm{b \perp}}(\omega,\alpha) &= \alpha \pm 90^{\circ}.
\label{eq:gamma-perp}
\end{align}
The first corresponds to the case when the spin velocity is directed, on the
$xy$ plane in the same direction of the polarization angle $\alpha$ of the
incoming beam; the second one corresponds to the case when the spin velocity is
directed perpendicularly respect to the polarization angle of the incoming
beam.

@@

Analyzing now the SVI fixing the velocity along the $\mathrm{a}$ Cartesian
coordinate on the surface of the structures we define the corresponding
magnitude as
\begin{align}
|\mathcal{V}^{\mathrm{a}}&(\omega,\alpha)| = \nonumber \\
&\sqrt { 
[\mathcal{V}^{\mathrm{ax}}(\omega,\alpha)]^{2} +
[\mathcal{V}^{\mathrm{ay}}(\omega,\alpha)]^{2} +
[\mathcal{V}^{\mathrm{az}}(\omega,\alpha)]^{2} .
},
\label{eq:vv-mag}
\end{align}
{\color{red} Then, the spin direction depends of the $x$, $y$, and $z$
components of the previous equation and so we define the spin orientation polar
and azimuthal angles as}
\begin{align}
\theta_{\mathrm{a}}  (\omega,\alpha)
=& 
\cos^{-1} \left( \frac{\mathcal{V}^{\mathrm{az}}(\omega,\alpha)}
{|\mathcal{V}^{\mathrm{a}}(\omega,\alpha)|} \right),
& 0 \leq &\theta \leq \pi, 
\label{eq:polar-ang}
\\
\varphi_{\mathrm{a}} (\omega,\alpha)
=& 
\tan^{-1} \left( \frac{\mathcal{V}^{\mathrm{ay}}(\omega,\alpha)}
{\mathcal{V}^{\mathrm{ax}}(\omega,\alpha)} \right),
& 0 \leq &\varphi \leq 2\pi.
\label{eq:azimuthal-ang} 
\end{align}
{\color{red} where $\theta_{\mathrm{a}}(\omega,\alpha)$ is measured from the
positive to the negative $z$ Cartesian coordinate and
$\varphi_{\mathrm{a}}(\omega,\alpha)$ is measured on the $xy$ plane in the
counter-clockwise direction from the positive $x$ Cartesian coordinate.}

@@

    \caption{Comparison of most intense responses of
    $\mathcal{V}^{\mathrm{ab}}$ for 2D \emph{alt} and \emph{up}, and bulk CdSe
    and GaAs structures and the corresponding polarization angles.}
    \label{fig:vab-str-comp}
\end{figure}

>>>>

\usepackage{amsmath}    % need for subequations
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs 
\usepackage{blindtext}  % fill text
\usepackage[dvipsnames]{xcolor}
\usepackage{multirow}
\usepackage[capitalize]{cleveref}
% \usepackage{showframe}
\usepackage[inline]{showlabels}

@@

Spintronics is an emerging research field of electronics in which the
manipulation and transport of spin of electrons in a solid state 
media plays the determining role adding a new degree of freedom to the
conventional charge manipulation.\cite{wolfSC04,fabianAPS07}
% 
At present there is an increasing interest in attain the same level of control
over the transport of spin at micro or nano scales as has been done for the
flow of charge in typical electronic devices.\cite{awschalomNP2007} Some
semiconductor spintronics devices have been proposed \cite{majumdarAPL06,
dattaAPL90,gotteNat16,pershinPRB08} and some of them require spin polarized
electrical current \cite{awschalomSSBM13} or pure spin current (PSC).
% 
One of the difficulties to achieve the development of spin current and PSC
semiconductor devices is the fact that the spin relaxation time in a
semiconducting media is short disabling the spin transport and then resulting
in a no observable spin current.\cite{murakamiSc03}
% 
In PSCs there is no net
motion of charge; spin-up electrons move in a given direction while spin-down
electrons travel in the opposite one. This phenomena can result from spin
injection,\cite{malPRB03} Hall Effects,\cite{sinovaPRB04} interference of two
optical beams,\cite{bhatPRL00, najmaiePRB03} or one photon absorption of
linearly polarized light\cite{bhatPRL05} and has been observed in gallium
arsenide (GaAs),\cite{zhaoPRL2006, stevensPRL03} aluminum-gallium arsenide
(AlGaAs),\cite{stevensPRL03} and Co$_2$FeSi.\cite{kimuraNGPAM12}

Graphene, an allotrope of carbon with hexagonal 2D lattice structure presents
properties like fractional quantum Hall effect at room temperature, excellent
thermal transport properties, excellent conductivity\cite{heerscheNat07} and
strength \cite{geimNM07, reinaNL08, novoselov2S07, balandinNL08} being then a
perfect platform to be used in two-dimensions electronic systems; however most
electronic applications are disabled by the absence of a semiconducting gap.
Recent studies demonstrate that the band gap of graphene can be opened by
applying an electric field,\cite{zhangN09} reducing the surface
area,\cite{hanPRL07} or applying uniaxial strain.\cite{niACSN08} Another
possibility to open the gap is by doping; this has been successfully achieved
using nitrogen,\cite{weiNL2009} boron-nitrogen,\cite{guoIJ11}
silicon,\cite{colettiPRB10} noble-metals,\cite{varykhalovPRB10} and
hydrogen.\cite{eliasS09, guisingerNL09, samarakoonACSN10}
% 
Depending on the percentage of hydrogenation and spatial configurations of
hydrogen-carbon bonds, hydrogenated graphene can result in different spatial
configurations.
% 
In this paper we present two 50\% hydrogenated graphene noncentrosymmetric
structures both presenting a discernible band gap: the \emph{up} structure,
shown in Fig. \ref{fig:up-struc}, has hydrogen atoms bonded to the carbon layer
only in the upper side of the structure while the \emph{alt} structure, shown
in Fig. \ref{fig:alt-struc}, has hydrogen alternating in the upper and bottom
sides of the carbon slab.\cite{zapataPSB2016}

Using those structures we address a theoretical study of the spin velocity
injection (SVI) by one-photon absorption of linearly polarized light.
% 
% We calculated the responses for the particular cases when the spin of
% electrons is directed along the the $z$ Cartesian coordinate, perpendicular
% to the $xy$ plane of the structure, or for the case when the velocity is
% directed along the $x$ and $y$ Cartesian direction on the $xy$ plane of the
% structures.
{\color{red}
% 
Because we have 2D structures we made the analysis for tow cases. The first is
fixing the spin of the electrons along the $z$ Cartesian direction with the
velocity directed on the surface of the structure in the $xy$ plane. The second
is fixing the spin velocity in the $x$ or $y$ direction and the spin directed
in $xyz$.}
% 
The SVI is an optical effect that quantifies the velocity at which a PSC moves
along the Cartesian direction $\mathrm{a}$ with the spin of electron polarized
along the Cartesian direction $\mathrm{b}$. One photon absorption of linearly
polarized light can promote a distribution of electrons in $\mathbf{k}$ space
regardless the symmetry of the material resulting in a null electrical current.
Then, the electrons excited to the conduction bands at opposite $\mathbf{k}$
points will result in opposite spin polarizations producing no net spin
injection.\cite{bhatPRL05} If the crystalline structure of the material is
noncentrosymmetric the spin polarization injected at a given $\mathbf{k}$ point
could not vanish\cite{alvaradoPRL85, schmiedeskampPRL88} 
% 
{\color{red}
% 
and then, since the velocities of electrons at opposite $\mathbf{k}$ points are
opposite, a pure spin current will be produced. Since the structures presented
here are noncentrosymmetric, they are good candidates in which this effect can
be induced.}

This paper is organized as follows. In Section \ref{sec:theory} we present the
theory and formulas that describe PSC and SVI. In Section \ref{sec:results} we
describe the details of calculations and the corresponding SVI spectra for the
\emph{up} and \emph{alt} structures. Finally, we present our conclusions in
Section \ref{sec:conclusions}.

@@

In this section, we report a summary of the theory that involves the PSC
phenomena from which rises the SVI treated in this paper.

In PSCs there is no net motion of electrical charge and spin-up electrons move
in a given direction while spin-down electrons travel in the opposite one.
This effect can result from one photon absorption of linearly polarized light
by a semiconductor, with filled valence bands and empty conduction bands,
illuminated by light with photon energy larger than the energy gap.
% note 1
Using a single continuous linearly polarized laser beam, it is possible
to promote electrons in $\mathbf{k}$ space regardless the symmetry of the
system resulting in a net current equal to zero. 
% 
If the phenomena is produced in a noncentrosymmetric semiconducting media, the
electrons promoted to the conduction bands at opposite $\mathbf{k}$ points
can produce a net spin polarization different to zero\cite{alvaradoPRL85},
resulting in a PSC because the velocity of electrons
at opposite $\mathbf{k}$ points are in opposite directions.

Using the multiple scale approach to solve the equation of motion for the
single particle density matrix $\rho_{mn}(\mathbf{k};t)$ we have that 
\begin{equation*}
\begin{aligned}
&\frac{\partial \rho_{cc'}}{\partial t} =
-i (\omega_{cc'} - i \epsilon) \rho_{cc'}
+
\frac{e^{2}E^{\mathrm{a}}(\omega)E^{\mathrm{b*}}(\omega)}
{i \hbar^{2}}
\times \\
& \quad \sum_{v}r^{\mathrm{a}}_{cv} r^{\mathrm{b}}_{vc'}
\left( \frac{1}{\omega - \omega_{c'v} - i \epsilon} 
- 
\frac{1}{\omega - \omega_{cv} + i \epsilon} \right).
\end{aligned}
\end{equation*}
Taking in the first term $\epsilon \rightarrow 0$ and changing the density
matrix operator to the interaction representation, $\tilde{\hat{\rho}} = e^{iH_
{0}t/\hbar} \hat{\rho} e^{-iH_{0}t/\hbar}$, with $H_{0}$ the ground state
Hamiltonian, the matrix elements can be written as
\begin{equation*}
\tilde{\rho}_{cc'(\mathbf{k})} 
= \langle c\mathbf{k}| \tilde{\hat{\rho}} | c'\mathbf{k} \rangle
= e^{i \omega_{cc'}t} \rho_{cc'}(\mathbf{k}).
\end{equation*}
where $H_{0}|n\mathbf{k}\rangle = \hbar \omega_{n}(\mathbf{k})|n
\mathbf{k}\rangle$ with $\hbar \omega_{n}(\mathbf{k})$ the energy of the
electronic band $n$ at point $\mathbf{k}$ in the irreducible Brillouin zone
(IBZ), and $|n\mathbf{k}\rangle$ the Bloch state. Then, the time derivative of
$\tilde{\rho}_{cc'}(\mathbf{k})$ is given by 
\begin{equation}\label{eq:derden}
\begin{aligned}
&\frac{d\tilde{\rho}_{cc'}(\mathbf{k})}{dt} 
= 
\frac{e^{2} E^\mathrm{a}(\omega) E^{\mathrm{b*}}(\omega)}{i\hbar^{2}} 
e^{i\omega_{cc'}t} 
\times \\
&\quad \sum_{v}r^{\mathrm{a}}_{cv} r^{\mathrm{b}}_{vc'}
\left( \frac{1}{\omega - \omega_{c'v} - i \epsilon} 
- 
\frac{1}{\omega - \omega_{cv} + i \epsilon} \right),
\end{aligned}
\end{equation}
where $\epsilon \rightarrow 0$ still needs to be taken. 
% 
The expectation value of an observable $\mathcal{O}$ with quantum mechanical
operator associated $\hat{\mathcal{O}}$ is given by
\begin{equation}\label{eq:trace}
\mathcal{O} = \mathrm{Tr}(\hat{\rho}\hat{\mathcal{O}}),
\end{equation}
where $\mathrm{Tr}$ denotes the trace. Then
\begin{align}
\mathcal{O} 
=&
\int \frac{d^{3}k}{8\pi^{3}}\sum_{c} \langle c \mathbf{k} 
| \hat{\rho} \hat{\mathcal{O}} |
c\mathbf{k} \rangle \nonumber \\
=& 
\int \frac{d^{3}k}{8\pi^{3}} \sum_{cc'} \tilde{\rho}_{cc'}(\mathbf{k}) 
\mathcal{O}_{c'c}(\mathbf{k}),
\label{eq:expectation}
\end{align}
where we used the interaction picture representation,
$\tilde{\mathcal{O}}_{cc'} = \langle c'\mathbf{k} | e^{iH_{0}t/\hbar}
\hat{\mathcal{O}} e^{-iH_{0}t/\hbar} | c\mathbf{k} \rangle$, and the closure
relationship $\sum_{c}|c\mathbf{k}\rangle \langle c\mathbf{k} = 1$. So, the
rate of change of $\mathcal{O}$ can be written from Eq. \eqref{eq:derden}, Eq.
\eqref{eq:trace}, and Eq. \eqref{eq:expectation} as
\begin{align}
\dot{\mathcal{O}} 
&=
\mathrm{Tr} \left( \frac{d \hat{\tilde{\rho}}}{dt} 
\tilde{\mathcal{O}} \right), 
\nonumber \\
&=\frac{e^{2}}{i\hbar^{2}} \int \frac{d^{3}k}{8\pi^{3}} 
\sum \mathcal{O}_{c'c} r^{\mathrm{a}}_{cv} r^{\mathrm{b}}_{vc'} \times
\nonumber \\
& \left( \frac{1}{\omega - \omega_{c'v} - i\epsilon} - 
\frac{1}{\omega - \omega_{cv} + i\epsilon} \right)
E^{\mathrm{a}}(\omega) E^{\mathrm{b*}}(\omega)
\label{eq:dotO}
\end{align}

Replacing the operator $\mathcal{O} \rightarrow \hat{K}^{\mathrm{ab}} \equiv v^
{\mathrm{a}} S^{\mathrm{b}}$ the matrix elements are given by
\begin{equation}
K^{\mathrm{ab}}_{nm}(\mathbf{k}) = 
\sum_{l} v^{\mathrm{a}}_{nl}(\mathbf{k})S^{\mathrm{b}}_{lm}(\mathbf{k}), 
\label{eq:velspimatelem}
\end{equation}
and using time-reversal invariance it follows that 
\begin{equation}
K^{\mathrm{ab}}_{nm}(\mathbf{-k}) = K^{\mathrm{ab*}}_{nm}(\mathbf{k}). 
\label{eq:k-timerev}
\end{equation}
Now the time derivative $\dot{K}^\mathrm{ab}$ can be obtained as we proceeded
in Eq. \eqref{eq:dotk-dev} of Appendix \ref{sec:mathder} and then we write
\begin{equation}
\dot{K}^{\mathrm{ab}}(\omega) =
\mu^{\mathrm{abcd}}(\omega)
E^{\mathrm{c}}(\omega) E^{\mathrm{d*}}(\omega),
\label{eq:dotk}
\end{equation}
where the pure spin-current pseudotensor components
\begin{equation}\label{eq:mu}
\begin{aligned}
\mu^{\mathrm{abcd}}  (\omega) &
=
\frac{\pi e^{2}}{\hbar^{2}} \int 
\frac{d^{3}K}{8 \pi^{3}} \times \\
\sum_{vcc'}^{'}
\mathrm{Re} & \left[ K^{\mathrm{ab}}_{cc'} 
\left( 
r^{\mathrm{c}}_{vc'} 
r^{\mathrm{d}}_{cv } +
(c \leftrightarrow d)
\right) 
\right] \delta(\omega-\omega_{cv}),
\end{aligned}
\end{equation}
are obtained from Eq. \eqref{eq:k-final} of the Appendix.
% 
The $'$ in the sum means that $c$ and $c'$ are quasi degenerate states and the
sum only covers these states and since $\mu^{\mathrm{abcd}}(\omega)$ is real we
have that $\mu^{\mathrm{abcd}}(\omega) = \mu^{\mathrm{abdc}}(\omega)$. This
previous equation is the same as Eq. (3) of Bhat et al.\cite{bhatPRL05}
obtained by using the semiconductor optical Bloch equations.

@@

We define the SVI as the velocity at which the spin, polarized along the
direction $\mathrm{a}$, propagates along the direction $\mathrm{b}$ as
\begin{equation}\label{eq:vab-w}
\mathcal{V}^{\mathrm{ab}}(\omega) \equiv
\frac{\dot{K}^{\mathrm{ab}}(\omega)}{(\hbar/2) \dot{n}(\omega)},
\end{equation}  
where $\dot{K}^{\mathrm{ab}} (\omega)$ is given by Eqns. \eqref{eq:dotk} and
\eqref{eq:mu} and the carrier
injection rate,
$\dot{n}(\omega)$, are given by
% 
\begin{equation}
\dot{n}(\omega) =
\xi^{\mathrm{ab}}(\omega) E^{c }(\omega) E^{d*}(\omega),
\label{eq:dotn}
\end{equation}
where $\xi^{\mathrm{ab}}(\omega)$ are the carrier
generation rate tensor components. 
% 
Since we have 2D structures we use an incoming electric field parallel to the
surface, $\mathbf{E}^{\mathrm{a}}(\omega,\alpha) = E^{\mathrm{a}}(\omega)
e^{i\alpha}$, where $\alpha$ corresponds to the linear polarization angle.
% 
Then from Eq. \eqref{eq:dotk}, Eq. \eqref{eq:vab-w}, and \eqref{eq:dotn}, and
including the polarization angle dependence we obtain
\begin{widetext}
\begin{align}
\mathcal{V}^{\mathrm{ab}}(\omega,\alpha)
% &= 
% \frac{2}{\hbar}
% \frac{\mu^{\mathrm{abxx}}(\omega)
% E^{2}(\omega)\cos^{2}(\alpha) + 
% \mu^{\mathrm{abyy}}(\omega)
% E^{2}(\omega)\sin^{2}(\alpha) + 
% 2\mu^{\mathrm{abxy}}(\omega)
% E^{2}(\omega)\cos(\alpha)\sin(\alpha)}
% {\xi^{\mathrm{xx}}(\omega)
% E^{2}(\omega)\cos^{2}(\alpha) + 
% \xi^{\mathrm{yy}}(\omega)
% E^{2}(\omega)\sin^{2}(\alpha)},
% \nonumber \\
&= 
\frac{2}{\hbar}
\frac{\mu^{\mathrm{abxx}}(\omega)\cos^{2}(\alpha) + 
\mu^{\mathrm{abyy}}(\omega)\sin^{2}(\alpha) + 
\mu^{\mathrm{abxy}}(\omega)\sin(2\alpha)}
{\xi^{\mathrm{xx}}(\omega)\cos^{2}(\alpha) + 
\xi^{\mathrm{yy}}(\omega)\sin^{2}(\alpha)},
\label{eq:vab-aw}
\end{align}
\end{widetext}
% 
Two interesting possibilities to analyze the SVI are fixing the spin along $z$,
directed perpendicularly to the surface of the structure, or fixing the
velocity along $x$ or $y$ on the $xy$ plane of the structures. Also we can
analyze the SVI contribution coming from each layer of the structure. In
following subsections we present these three cases.

@@

Analyzing the SVI, Eq. \eqref{eq:vab-aw}, we define the magnitude of the spin
velocity with spin polarized along the $\mathrm{b}$ direction as
\begin{equation}
|\mathcal{V}_{\sigma^{\mathrm{b}}}(\omega,\alpha)| 
=
\sqrt{
[\mathcal{V}^{\mathrm{xb}}(\omega,\alpha)]^{2}\ +
[\mathcal{V}^{\mathrm{yb}}(\omega,\alpha)]^{2}\ 
}, 
\label{eq:vs-mag}
\end{equation}
and the angle at which the spin velocity is directed on the $xy$ plane as
\begin{equation}
\gamma_{\mathrm{b}} (\omega,\alpha)
=
\tan^{-1} \left( \frac{\mathcal{V}^{\mathrm{yb}}(\omega,\alpha)}
{\mathcal{V}^{\mathrm{xb}}(\omega,\alpha)} \right),
\label{eq:gamma-ang}
\end{equation}
where this angle is measured in the counter-clockwise direction from the
positive $x$. We also define two special angles
\begin{equation}
\gamma_{\mathrm{b \parallel}}(\omega,\alpha) = \alpha, 
\label{eq:gamma-par} 
\end{equation}
and
\begin{equation}
\gamma_{\mathrm{b \perp}}(\omega,\alpha) = \alpha \pm 90^{\circ}.
\label{eq:gamma-perp}
\end{equation}
The first corresponds to the case when the spin velocity is directed, in the
same direction of the polarization angle $\alpha$ of the incoming beam and the
second one corresponds to the case when the spin velocity is directed
perpendicularly with respect to the polarization angle of the incoming beam.

@@

Fixing the velocity along $x$ or $y$ we define the corresponding magnitude as
\begin{align}
|\mathcal{V}^{\mathrm{a}}&(\omega,\alpha)| = \nonumber \\
&\sqrt { 
[\mathcal{V}^{\mathrm{ax}}(\omega,\alpha)]^{2} +
[\mathcal{V}^{\mathrm{ay}}(\omega,\alpha)]^{2} +
[\mathcal{V}^{\mathrm{az}}(\omega,\alpha)]^{2} .
},
\label{eq:vv-mag}
\end{align}
Then, the spin direction depends on the components of the previous equation and
so we define the spin orientation polar and azimuthal angles as
\begin{align}
\theta_{\mathrm{a}}  (\omega,\alpha)
=& 
\cos^{-1} \left( \frac{\mathcal{V}^{\mathrm{az}}(\omega,\alpha)}
{|\mathcal{V}^{\mathrm{a}}(\omega,\alpha)|} \right),
& 0 \leq &\theta \leq \pi, 
\label{eq:polar-ang}
\\
\varphi_{\mathrm{a}} (\omega,\alpha)
=& 
\tan^{-1} \left( \frac{\mathcal{V}^{\mathrm{ay}}(\omega,\alpha)}
{\mathcal{V}^{\mathrm{ax}}(\omega,\alpha)} \right),
& 0 \leq &\varphi \leq 2\pi.
\label{eq:azimuthal-ang} 
\end{align}
% where $\theta_{\mathrm{a}}(\omega,\alpha)$ is measured from the
% positive to the negative $z$ Cartesian coordinate and
% $\varphi_{\mathrm{a}}(\omega,\alpha)$ is measured on the $xy$ plane in the
% counter-clockwise direction from the positive $x$ Cartesian coordinate.

@@

    \caption{Comparison of most intense responses of
    $\mathcal{V}^{\mathrm{ab}}$ for 2D \emph{alt} and \emph{up}, and bulk CdSe
    and GaAs structures and their corresponding polarization angle $\alpha$.}
    \label{fig:vab-str-comp}
\end{figure}
